\documentclass{article}

% natbib options must be passed before neurips_2025 loads it.
\PassOptionsToPackage{numbers,sort&compress}{natbib}

% "preprint" = publicly posted, non-anonymous version with authors shown.
% (Use \usepackage{neurips_2025} for the anonymous double-blind submission, or
%  \usepackage[final]{neurips_2025} for the camera-ready.)
\usepackage[preprint]{neurips_2025}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{siunitx}
\usepackage{enumitem}
\usepackage{placeins}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,backgrounds,calc,shapes.geometric}

% Figures (PNGs + the TikZ \input diagrams) live in figures/.
\graphicspath{{figures/}{./figures/}}

% Convenience macros used by the section sources.
\newcommand{\degr}{\ensuremath{^\circ}}
\newcommand{\fd}{\ensuremath{f_d}}
\newcommand{\theref}{\ensuremath{\theta}}

\title{Automated Ultrasound Doppler Angle Estimation Using Deep Learning}

\author{%
  Nilesh Patil \\
  Goergen Institute for Data Science\\
  University of Rochester Medical Center\\
  University of Rochester\\
  Rochester, NY, USA \\
  \And
  Ajay Anand \\
  Goergen Institute for Data Science\\
  University of Rochester\\
  Rochester, NY, USA \\
}

\begin{document}
\maketitle

\begin{abstract}
The Doppler angle $\theta$, measured between the ultrasound beam and the
blood-flow direction, enters the Doppler equation through $\cos\theta$, so an
error in $\theta$ propagates into the estimated blood-flow velocity, and the
propagation is severe as $\theta$ approaches \SI{90}{\degree}. Patil and Anand
(EMBC 2019) proposed estimating this angle from B-mode carotid images with a
deep convolutional network. We replicate that pipeline and isolate the design
choice it depends on. Standard global average pooling is approximately
rotation-invariant and discards the orientation signal the target encodes; an
orientation-preserving grid-pooling head lifts a frozen DenseNet201 backbone to
\SI{5.84}{\percent} mean absolute percentage error (MAPE). We then tune the
regression head with the Optuna TPE sampler and ensemble five backbones. We
report pooled out-of-fold metrics under two sampling protocols. Image-level
sampling over the rotation-augmented corpus reaches \SI{2.79}{\percent} MAPE
(\SI{1.96}{\degree} MAE, $R^2=0.995$); patient-level sampling, which holds out
whole volunteers, reaches \SI{8.53}{\percent} MAPE (\SI{5.93}{\degree} MAE,
$R^2=0.952$). On this small, narrow-domain task DenseNet201 outperforms
ConvNeXt and EfficientNet backbones. Split-conformal \SI{90}{\percent}
prediction intervals span $\pm$\SI{20.5}{\degree}. Bland--Altman analysis
against the single reference reading gives a \SI{-4.3}{\degree} bias with
\SI{95}{\percent} limits of agreement $-24.25$ to $+\SI{15.63}{\degree}$
(method-vs-reference, not inter-observer). Rotation test-time augmentation
roughly halves the base-image error, and circular fusion with a structure-tensor
angle prior reaches \SI{2.72}{\degree} MAE. Tables and figures are regenerated
from code.
\end{abstract}

\input{sections/intro}
\FloatBarrier
\input{sections/methods}
\FloatBarrier
\input{sections/results}
\FloatBarrier
\input{sections/evaluation}
\FloatBarrier
\input{sections/discussion}
\FloatBarrier

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
